Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-15 analysis/paper/paper.qmd
Chose Ben's edits instead of GP's edits

The onset of the Late Paleolithic period in Korea is marked by the appearance of stemmed points around 40-35 ka. The stemmed point is a projectile point made out of an elongated flake or blade with slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft. The stemmed point was the first composite tool type appearing on the Korean Peninsula, and represents new hunting strategies there, as well as other adjacent regions where it appeared later [@seong2008tanged; @lee2019were]. Stemmed points, combined with blade technology, multi-stage production sequences, and evidence of being resharpened and reused, are important aspects of the Late Paleolithic technological innovations in Korea because these features are rarely seen in earlier periods [@chang2013human; @bamforth2009projectile; @seong2015diversity]. Currently the earliest stemmed points in Northeast Asia are from Yongho-doing site in Korea dated to 38.5ka and made on elongate flakes [@seong2009emergence; @seong2015diversity; @bae2017origin]. After their first appearance in Korea, the stemmed points spread to the Japanese archipelago [@chang2013human].

@@

Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two of these biases can be identified in stone artifact assemblages. Their research focused on metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600. They equated guided transmission (where individuals learn new behaviors through trial and error) with high metric variation and low correlation between metric attributes. Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables. They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools. From these results, they inferred a social context in eastern California of distant and unfamiliar neighbors, with limited contact between groups to acquire technology directly. In this context of limited contact, they argue that the new technologies developed largely by trial and error. On the other hand, in central Nevada they found that metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and lengths on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of the archaeological projectile points are closest to the metrics of the simulated assemblage with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work shows how the cultural transmission of tool-making behaviours can be measured, simulated, and used to interpret the archaeological record.

Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information of the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of the points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are important because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). If this social context was prevalent, we predict this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with other groups. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, the model, or ideal, stemmed point is likely to have been a highly successful and frequently chosen one, and learners copy all information about the point design as a package. This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf. Bae's migration model), where stemmed points first appeared and then spread from. In this social context, indirect bias is the dominant influence on transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points should be more standardized, and attributes on the stemmed points will be more correlated, and assemblages from multiple sites will show less variance between and within them.

@@

After the first discovery of stemmed points at the Seokjangri site in the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map). Most sites contain only a few points but a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that are unbroken from the tip to the stem. We excluded artifacts that were recorded as stemmed points but lack a stem. This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number). The images of the stemmed points were obtained from published excavation reports and by direct photography during our research on local museum collections. We defined multiple assemblages in a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were built on radiocarbon ages, and for sites without any, the presence/absence of blades, stemmed point blanks, and toolkit composition were used to assign assemblages to a phase. Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blade in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points in into these three phases to facilitate observations of change over time.

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs. Shale is the dominant raw material in both phase 2 and 3. In phase 3 there is an increase in the proportions of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B in @fig-raw-materials-by-phase shows the count of stemmed points by raw material type at assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.

@@

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we used morphological attributes on the stemmed points to examine variations in shape and the relationship between each attribute (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yields more easily interpretable numerical and visual outcomes than traditional caliper measurements [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness as these measurements were not available to us. For the landmark analysis, we put a total of 11 landmarks on the outline of each stemmed point and computed distances between pairs of landmark coordinates to derive attributes for our statistical analyses. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. We used the point tool in ImageJ [@schneider2012nih] to capture the landmarks from images of the artifacts, and export them as XY coordinate data for further analyses.

@@

Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points. The correlation of artifact attributes was proposed by @bettinger1999point to be a key indicator of indirect bias during cultural transmission. This is based on the assumption that artifact-making was transmitted in packages of traits taken from socially successful individuals who were role-models. In their work, @bettinger1999point interpreted correlation coefficient values of around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation. In this research, we examine correlation between attributes and explore other variables such as different phases, raw materials and assemblages.

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is the ratio (usually expressed as a percentage) of the standard deviation to the mean. It has been used as a robust statistical method to calculate the degree of standardization, precision, equality, homogeneity, etc. in various disciplines [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between different types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template. Among the various artifacts for which \@eerkens2001techniques summarised CV values, the most comparable to stemmed points are Great Basin projectiles points, whose CV values range from 6-55%, with an average of 22%. Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].

Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values has varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements when using small samples. For example, @vanpool2011quantitative proposed a "corrected CV", for a smaller samples (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, most of our analyses here involves comparisons of smaller subsets of this sample. For example to explore temporal and regional patterns we compared samples of well under 25 pieces. Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use. We also computed confidence intervals using method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].

One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases. @bettinger1999point did not identify a threshold for CV values. Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases. We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias. With the exception of weight, the CVs for metric attributes on the Rosegate points is 17% to 24%. We believe higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on following reasons. First, variation is generated by small errors that are transmitted between individuals and the errors gets bigger through generations [@eerkens2005cultural]. The duration of transmission process for stemmed points is much longer (in the order of thousands of years) than the case of Rosegate (in the order of hundreds of years in Fremont region) [@bischoff2021rosegate]. Second, the errors are likely to vary by raw materials. Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values than compared to standard CV values [@vanpool2011quantitative]. Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likley reflect indirect bias.

@@

Panel D of @fig-corr-analysis shows that the correlations between attributes became stronger over time. The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold. value. <!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.

@@

Our correlation analyses show that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage). When all artefacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. However, further analysis by site and raw material show more ambiguous patterns, with CV values close to or below the threshold value. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological phase 3, after 25 ka. This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also confirm a slight decrease in CV values in four attributes. This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model. Drawing on evolutionary ecological theory, we might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.

In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages. Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase 3. Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Yonsangdong that date to this period. One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.

>>>>

The emergence of stemmed points marks the beginning of the Late Paleolithic period in Korea around 40-35 ka. A stemmed point is a projectile point made out of an elongated flake or blade with a slight retouch on the proximal end to shape an acute tip, and on the distal end to make a stem, which connects to a shaft. Stemmed points were the first composite tool types to appear on the Korean Peninsula, and were a symbol of new hunting strategies there and in adjacent regions [@seong2008tanged; @lee2019were]. A number of technological innovations are evident on stemmed points, including blade technology, multiple manufacturing stages, and evidence of resharpening and reusing, which were rarely seen earlier [@chang2013human; @bamforth2009projectile; @seong2015diversity]. Currently, the oldest stemmed points in Northeast Asia are from the Yongho-doing site in Korea, dating back to 38.5ka [@seong2009emergence; @seong2015diversity; @bae2017origin]. Following their appearance in Korea, stemmed points spread to the Japanese archipelago [@chang2013human].

@@

Bettinger and Eerkens [-@bettinger1997evolutionary; -@bettinger1999point; -@eerkens2008cultural] have shown how two of these biases can be identified in stone artifact assemblages. Their research focused on the metric variables of stone points during the introduction of bow and arrow technology in the Great Basin around AD 300-600. They equated guided transmission (where individuals learn about new behaviors through trial and error) with high metric variation and low correlation between metric attributes. Conversely, they inferred indirect bias (where a behavior is transmitted because of its association with other attributes, such as the prestige or skill of other individuals) from less variation and more correlated variables. They found that artifacts in the new bow and arrow technology in eastern California have low correlations of basal width and mass, which they interpreted as a result of guided transmission dominating the introduction and spread of these tools. Based on these results, they inferred that eastern California had a social context of distant and unfamiliar neighbors with little direct contact between groups. In this context of limited contact, they argue that the new technologies developed largely by trial and error. On the other hand, in central Nevada they found that metric variables are highly correlated, indicating that the bow and arrow was introduced and spread by indirect bias.

Following Bettinger and Eerkens' framework, Garvey [-@garvey2018current] uses simulation to explore the degree of standardization represented by coefficient of variation (CV) values of projectile points from the US Southwest and westernmost southern High Plains. Garvey measured CV values for weight, thickness, width and length on two types of projectile points, Washita and Fresno points, from the Henderson site. She then computed simulated CV values according to three scenarios of different levels of transmission fidelity (i.e. CV = 10%, 5%, and 3%). The observed metrics of archaeological projectile points are closest to simulated metrics with 3% CV, which represents "extremely high-fidelity copying". A CV of 3% is the Weber fraction, or threshold of human visual perception, with variation below this value being too subtle for people to notice [@eerkens2000practice]. Garvey's work demonstrates how the cultural transmission of tool-making behaviors can be measured, simulated, and interpreted from the archaeological record.

Garvey and Bettinger and Eerkens inferred transmission processes from material culture because they were unable to observe them directly. Direct observations have been reported by Mesoudi and O'Brien [-@mesoudi2008cultural] who conducted experimental research using groups of undergraduate students to test the underlying assumptions of Bettinger and Eerkens' study. They simulated model-based bias in projectile point making by providing the design of a model and information about the model's prior success to their research participants. They also simulated guided variation by allowing their participants to explore their own designs. They observed that the majority of participants who were able to choose the previous design copied the most-successful model. Metric attributes of projectile points made when copied from successful models were more highly correlated than attributes of points made by trial and error. Mesoudi and O'Brien's experimental results are critical because they confirm the robustness of previous assumptions about cultural transmission biases, and validate the connection between material culture variation and cultural transmission biases.

@@

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement (cf. Seong's *in situ* model). If this social context was prevalent, we predict that this trial and error behavior could have left a distinctive signature on the metric variables of the stemmed points. If trial and error was the dominant bias in the transmission of knowledge about how to make stemmed points, we expect that the morphological attributes on a stemmed point will be poorly correlated and show high variation within assemblages and between sites.

On the other end of our spectrum, we have socially connected groups that occupied places close to other groups or had regular contact with them. Relatively high degrees of social connectivity provide frequent opportunities for observing others and acquiring information. Individuals or groups learned the technology of stemmed points by copying a model from another individual or group. In this scenario, it is likely that the model, or ideal, stemmed point has been highly successful and frequently chosen, and learners copy all information about the point design as a package. This implies a small number of locations, or a single point (perhaps from a source external to the Korean peninsula, cf. Bae's migration model), where stemmed points first appeared and then spread from. In this social context, indirect bias is the dominant influence on the transmission of lithic technology. If indirect bias was prominent in the Korean Paleolithic, we expect that stemmed points would be more standardized, attributes on the stemmed points would be more correlated, and assemblages from multiple sites would exhibit less variance between and within them.

@@

After the first discovery of stemmed points at the Seokjangri site at the 1960s, around 300 have been found in nearly 30 sites across Korea (@fig-map). While most sites contain only a few points, a few sites have many more, such as Suyanggae (n = 55), Jingeuneul (n = 99), and Yongsandong (n = 38) [@Kim_2017]. Among these stemmed points, we selected those that are unbroken from the tip to the stem. We excluded artifacts that were recorded as stemmed points but lack a stem. This resulted in a sample of `r total_number_of_sp` stemmed points from `r total_number_of_assemblages` assemblages unearthed from `r total_number_of_sites` sites spanning the period 40-17 ka (@tbl-sp-number). The images of the stemmed points were obtained through published excavation reports and direct photography during our research in local museums. We defined multiple assemblages from a site where artifact-bearing deposits were separated by culturally sterile deposits, or where distinct artifact-bearing stratigraphic units could be identified by major differences in the texture, color, and composition of the sedimentary deposits.

@@

To analyze morphological change over time, we used previously developed chronologies that identify three phases in the Korean Late Paleolithic [@seong2015diversity; @Park_2013]. These chronologies were constructed based on radiocarbon ages, and assemblages without radiocarbon ages were classified by blades, stemmed point blanks, and toolkit composition. Following these previous chronological schemes, we divided the Korean Late Paleolithic assemblages into three chronological phases: 1) stemmed points made out of flakes and no blades in assemblages, 2) stemmed points made out of blades or flakes and the existence of blades in assemblages, and 3) stemmed points made out of blades and the existence of micro blades in assemblages (@tbl-korean-chronology). Applying this division is necessary because some sites such as Yonghodong, Goryeri, Jungmal, and Mungyeong have no radiocarbon ages due to research limitations. Inferring ages for these sites by analogy to the technological sequences at sites with dates is necessary for making maximal use of the available archaeological data to increase our sample size. We arranged our `r total_number_of_assemblages` assemblages containing stemmed points into these three phases to facilitate observations of change over time.

@@

Panel A of @fig-raw-materials-by-phase summarizes the amount of stemmed points in each chronological phase, and the distribution of raw materials across the phases. We exclude phase 1 from our analyses below because of the small sample size to compute correlations and CVs. Shale is the dominant raw material in both Phase 2 and 3. In Phase 3 there is an increase in the proportion of porphyry, driven largely by finds from Yongsangdong, and other raw materials. Panel B of @fig-raw-materials-by-phase shows the count of stemmed points by raw material type in assemblages with more than five stemmed points. Much of the raw material diversity comes from isolated finds, with those assemblages with more than five points showing high homogeneity in raw materials.

@@

Following the metric attributes used in previous studies of cultural transmission and projectile points, i.e. maximum length and width [@bettinger1997evolutionary; @bettinger1999point; @eerkens2008cultural; @garvey2018current], we examined variations in shape and the relationship between each attribute using morphological attributes on stemmed points (@fig-labelled-schematic). We obtained our morphological attribute data from landmark analysis of digitized images of stemmed points. Compared to traditional caliper measurements, morphometric data yield more easily interpretable numerical and visual outcomes [@buchanan2010geometric; @thulman2012discriminating; @cardillo2016environment; @petvrik2018shape; @macleod2018quantitative; @okumura2019archaeology; @suarez2019life]. We did not include weight and thickness since these measurements were not available to us. For the landmark analysis, we placed 11 landmarks on the outline of each stemmed point and calculated distances between landmark coordinate pairs to derive attributes for statistical analysis. The landmarks we recorded are described in @tbl-landmark-abbreviations and shown in @fig-labelled-schematic. Using the point tool in ImageJ [@schneider2012nih] we captured the landmarks from images of the artifacts, and exported them as XY coordinate data for further analyses.

@@

Following @bettinger1999point we computed Pearson's correlation coefficients for all attributes on the stemmed points. The correlation between artifact attributes was proposed by @bettinger1999point as a key indicator of indirect bias during cultural transmission. It is based on the assumption that artifact-making was transmitted in packages of traits inherited from socially successful individuals who were role models. In their work, @bettinger1999point interpreted correlation coefficient values around `r threshold_value_cor` and higher as evidence of indirect bias, and lower values as evidence of guided variation. In this research, we examine the correlation between attributes and explore other variables, such as different phases, raw materials and assemblages.

We chose Coefficients of Variation (CV) to measure the variation among the measurements of the stemmed points. The CV is determined by the ratio of the standard deviation to the mean (usually expressed as a percentage). The method has been used in various disciplines to calculate standardization, precision, equality, homogeneity, etc. [@ng2006performance; @panichkitkosolkul2009improved; @panichkitkosolkul2013confidence; @wang2020standardization]. In archaeological studies, CV has been applied to measure the variation between artifacts and to test hypotheses about cultural evolutionary processes, including distinguishing between the types of learning biases affecting cultural transmission [@eerkens2001techniques; @eerkens2005cultural; @schillinger2014copying; @garvey2018current]. As a guide to interpreting CV values, Eerkens and Bettinger [-@eerkens2001techniques] claim that over 57.7% of the CV is the result of random production and below 1.7% is the by-product of using a scale or template. Among all the various artifacts for which \@eerkens2001techniques summarized CV values, the most comparable to stemmed points are Great Basin projectile points, whose CV values range from 6-55%, with an average of 22%. Similarly, projectile points from the US Southwest have CV values ranging from 11% to 33% [@garvey2018current].

Sample size can impact the accuracy of CV estimates [@toebe2018sample; @kelley2007sample; @vanpool2011quantitative]. In previous archaeological studies, sample sizes for CV values have varied from five to more than a thousand artifacts [@bettinger1999point; @garvey2018current; @costin1995standardization; @wierer2013variability; @kvamme1996alternative; @rivals2009new]. A variety of methods are available for improving the reliability of CV measurements when using small samples. @vanpool2011quantitative, for instance, proposed a "corrected CV" for smaller samples (i.e. n\<25). Statistical research has resulted in several methods for computing confidence intervals on CVs to show uncertainty [@koopmans1964confidence; @miller1991asymptomatic; @sharma1994asymptotic; @mckay1932distribution; @vangel1996confidence; @curto2009coefficient; @banik2011estimating; @panichkitkosolkul2013confidence; @mahmoudvand2009two; @gulhar2012comparison]. Although we have a total of `r total_number_of_sp` artifacts, most of our analyses involve comparisons of smaller subsets. As an example, we compared samples of less than 25 pieces in order to explore temporal and regional patterns. Given these relatively small sample sizes in our study, we used the corrected CV formula, which we have implemented here in a function in the R programming language for others to use. We also computed confidence intervals using the method proposed by @sharma1994asymptotic to increase the credibility of the comparisons and be transparent about the precision and accuracy of our results [@kelley2007sample; @albatineh2014confidence].

One limitation of previous work is the absence of a clear threshold value for interpreting CV values in terms of different transmission biases. @bettinger1999point did not identify a threshold for CV values. Here we propose `r threshold_value` as an approximate threshold to distinguish between the transmission biases. We used data in @bettinger1999point to calculate CV values of Rosegate points from Monitor Valley, California, which was speculated by Bettinger and Eerkens to be the byproducts of indirect bias. With the exception of weight, the CVs for metric attributes on Rosegate points range from 17% to 24%. We believe the higher side of the CV range of Rosegate points is appropriate as a threshold value for our research based on the following reasons. First, variation is generated by small errors that are transmitted between individuals and the errors get bigger through generations [@eerkens2005cultural]. The duration of the transmission process for stemmed points is much longer (in the order of thousands of years) than in the case of Rosegate (in the order of hundreds of years in the Fremont region) [@bischoff2021rosegate]. Second, the errors are likely to vary by raw materials. Some raw materials, such as clay, are easier to control variation in while less controllable materials such as stone are likely to have higher CV values because of the relatively unpredictable nature of flaking [@eerkens2001techniques]. We assume that Rosegate projectile points were made from more finer grained raw materials such as flint compared to the raw materials for Korean stemmed points, which include shale, rhyolite, porphyry, etc. (@fig-raw-materials-by-phase). Third, the corrected CV calculation for small sample sizes that we are using tends to result in slightly higher values compared to standard CV values [@vanpool2011quantitative]. Given the tentative identification of this threshold value, we premised that CV values that are lower than `r threshold_value` likely represent guided variation while CV values over `r threshold_value` likely reflect indirect bias.

@@

Panel D of @fig-corr-analysis shows that the correlations between attributes became stronger over time. The median correlation value for Phase 2 is much lower than `r threshold_value_cor`, which @bettinger1999point interpreted as a threshold to distinguish between transmission biases, while the Phase 3 median is close to this threshold. value.

@@

Our correlation analyses show that the correlation coefficients of stemmed points are mostly positive and mostly at or under `r threshold_value_cor`, showing the possibility of guided variation as their main transmission bias. Some correlations such as body length and maximum length have strong relationships while other correlations such as tang length and body length are much weaker. The correlation coefficients increase from chronological Phases 2 to 3 (Panel D of @fig-corr-analysis), but when we look at individual assemblages the pattern is not consistent and the values vary depending on the assemblage (@fig-plot-cor-four-assemblage). When all artefacts are considered together, CV values for all stemmed point attributes are mostly over `r threshold_value`, except for maximum length, which is just below. However, further analysis by site and raw material show more ambiguous patterns, with CV values close to or below the threshold value. Change in CV values for over time in the Korean Late Paleolithic period is complex, with no clear directional changes (@fig-plot-cv-over-time). Comparing chronological Phases 2 and 3, we found that there are only minor differences between the two phases. Applying the Modified Signed-Likelihood Ratio Test (MSLR) to test for the equality of CVs [@smallwood2022using; @krishnamoorthy2014improved], we found no statistically significant changes in the CV values of any attributes (@tbl-cv-over-time) between the two phases. In exploring individual assemblages that contain multiple stemmed points we found minimal change in CV values over time (@fig-plot-cv-four-assemblage).

@@

<!-- GP need to add this thing in the discussion -->
<!-- this next sentence is probably best in the discussion section --> Based on this results, we assume that the earlier transmission for stemmed points was more likely the results of guided variation. On the other hand, Phase 3 could be more associated with indirect bias.
The results of this study have implications for determining whether these novel technologies originated outside of the Korean Peninsula or if they were locally developed independently.??

To contextualize this further, on one end of our spectrum we have socially isolated groups who stayed in physically remote places from other groups, or had unfamiliar neighbors with limited contact between groups. Our assumption for this social context is that individuals or groups acquired the technology of stemmed points by modifying existing flake tool forms through trial-and-error processes to solve problems relating to resource procurement.

@@

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological Phase 3, after 25 ka. This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also confirm a slight decrease in CV values in four attributes. This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model. Drawing on evolutionary ecological theory, we might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [cf. @fitzhugh2001risk]. With more frequent contacts and connections with members of social networks, there may have been more opportunities to learn point-making directly from socially successful individuals.

In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages. Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in Phase 3. Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in Phase 3, and dominating assemblages at Yonsangdong that date to this period. One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during Phase 3.
